---
title: "Predicting Insurance Cost"
subtitle: "Insurance Cost Analysis"
institute: "The Ohio State University"
author: "Ohenmaa Sarfo-Bonsu"
date: "2025-12-16"
execute:
  echo: true
knitr:
  opts_chunk:
    message: false
    warning: false
format:
  html:
    toc: true
embed-resources: true
output: pdf_document
---


```{r}
#dataset link: https://www.kaggle.com/datasets/mirichoi0218/insurance
# it was further appended using AI
library(MASS)
library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)
library(car)
```


# Load the Dataset
```{r}
# "insurance_extended.csv"
# "insurance_realistic_augmented.csv"
# "insurance.csv"
df <- read.csv("insurance.csv")
df
```

# Data Cleaning
```{r}
sum(is.na(df))
summary(df)
# Nothing looks off
```


# Exploratory Data Analysis


```{r}
ggpairs(
  df |> select(
   age, 
   bmi,
   charges
  ),
  columns = 1:3,
  aes(alpha = 0.6),
  upper = list(
    continuous = wrap(
      "cor",
      digits = 2,
      size = 2.2,
      mapping = aes()
    )
  ),
  lower = list(
    continuous = wrap("points", size = 0.5, color = "coral")
  ),
  diag = list(
    continuous = wrap("barDiag", bins = 20, fill = "coral")
  ),
  progress = FALSE
)+ labs(caption = "Figure 2: Scatterplot matrix bivariate and univariate distributions") +
   theme(
    plot.caption = element_text(hjust = 0)
  )

```
Charges and Age have a positive linear correlation. There is also a roughly positive linear relationship between BMI and charges, as well as Age and BMI. Age is roughly reminiscent of a uniform distribution, while the distribution closest to normal is BMI. Weak correlation observed between the predictors, which ensures that each predictor brings new information to the model. 

```{r}
df |>
  ggplot(aes(x = region, y= charges)) +
  geom_boxplot()

df |>
  ggplot(aes(x = smoker, y= charges)) +
  geom_boxplot()

df |>
  ggplot(aes(x = sex, y= charges)) +
  geom_boxplot()
```
The biggest discrepancy of mean insurance cost can be observed between smoking and non-smoking groups. 


# Multiple Linear Regression Model

```{r}
# split data into train and test

set.seed(123) # for reproducibility
train_index <- sample(1:nrow(df), 0.8 * nrow(df))
train_data <- df[train_index, ]
test_data <- df[-train_index, ]


# validating the split
nrow(train_data) / nrow(df)
nrow(test_data) / nrow(df)

```

```{r}
# run regression model
model <- lm(charges ~ (.)^2, data = train_data)
# also (.)^2 includes main effects AND all 2-way interactions 

# used to be model <- lm(charges ~ (.)^2, data = df) but i changed it becuase there's no point in splitting into a train/test if you're just going to turn around and use the entire dataframe df to train your model. 

# after getting our (.)^2 model, we perform stepwise model selection, picking a model based on the AIC. 

summary(model)

aic_model <- stepAIC(model, direction = "both")
summary(aic_model)
```


```{r}
# validate with test set


# 1. Get the actual values from your test set
actuals <- test_data$charges  # Replace with your actual column name

# 2. Get your predictions from your model (you've already done this)
predictions <- predict(aic_model, newdata = test_data)

# 3. Calculate Metrics
mse <- mean((actuals - predictions)^2) # mean squared error
rmse <- sqrt(mse)
mae <- mean(abs(actuals - predictions))
rss <- sum((actuals - predictions)^2)
tss <- sum((actuals - mean(actuals))^2)
r_squared <- 1 - (rss / tss) # r squared



# Calculate MAPE
mape <- mean(abs((actuals - predictions) / actuals)) * 100
cat("The average error is", mape, "%\n")
cat("The 'Accuracy' is", 100 - mape, "%\n")
r_squared

```

```{r}
#further model analysis

summary(aic_model)
plot(aic_model)
```










